Computing in the Field of Rationals

### 2012-06-21 rational_ext.tex
_

The semantics of the primitive combinators is given by the following
single-step reductions. Since there are no values of type {{0}}, the rules
omit the impossible cases:
\begin{scriptsize}
%subcode{opsem} include main
%! columnStyle = rlcl
% swap+ & (left v) &|-->& right v
% swap+ & (right v) &|-->& left v 
% assocl+ & (left v1) &|-->& left (left v1)
% assocl+ & (right (left v2)) &|-->& left (right v2)
% assocl+ & (right (right v3)) &|-->& right v3 
% assocr+ & (left (left v1)) &|-->& left v1
% assocr+ & (left (right v2)) &|-->& right (left v2)
% assocr+ & (right v3) &|-->& right (right v3)
% identl* & ((), v) &|-->& v 
% identr* & v &|-->& ((), v) 
% swap* & (v1, v2) &|-->& (v2, v1) 
% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
% dist & (left v1, v3) &|-->& left (v1, v3)
% dist & (right v2, v3) &|-->& right (v2, v3)
% factor & (left (v1, v3)) &|-->& (left v1, v3) 
% factor & (right (v2, v3)) &|-->& (right v2, v3)

@@

@@

\subsection{Small Step Semantics}

>>>>

The operational semantics of {{Pi}} is summarized below (also see
\cite{infeffects}).  The semantics of the primitive combinators is
given by the following single-step reductions below. Since there are
no values of type {{0}}, the rules omit the impossible cases:
\begin{scriptsize}
%subcode{opsem} include main
%! columnStyle = rlcl
% swap+ & (left v) &|-->& right v
% swap+ & (right v) &|-->& left v 
% assocl+ & (left v1) &|-->& left (left v1)
% assocl+ & (right (left v2)) &|-->& left (right v2)
% assocl+ & (right (right v3)) &|-->& right v3 
% assocr+ & (left (left v1)) &|-->& left v1
% assocr+ & (left (right v2)) &|-->& right (left v2)
% assocr+ & (right v3) &|-->& right (right v3)
% identl* & ((), v) &|-->& v 
% identr* & v &|-->& ((), v) 
% swap* & (v1, v2) &|-->& (v2, v1) 
% assocl* & (v1, (v2, v3)) &|-->& ((v1, v2), v3) 
% assocr* & ((v1, v2), v3) &|-->& (v1, (v2, v3)) 
% dist & (left v1, v3) &|-->& left (v1, v3)
% dist & (right v2, v3) &|-->& right (v2, v3)
% factor & (left (v1, v3)) &|-->& (left v1, v3) 
% factor & (right (v2, v3)) &|-->& (right v2, v3)

The operational semantics of the closure conditions are presented in
the usual big-step style.

@@

The type safety of {{Pi}} follows directly from the corresondence
between inductive definitions of the types and the big-step
semantics. Previous work also established the following properties:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Categorical Structure}

For the purpose of establishing equivalence of morphisms in the
categorical presentation we use the following notion of equality:

\begin{lemma}[{{c1  = c2}}]
We say {{c1 = c2}} if {{c1:b1<->b2}} and {{c2:b1<->b2}} and for all
{{v1:b}}, we have {{c1 v1 |--> v2}} iff {{c2 v1 |--> v2}} and {{v2:b2}}.
\end{lemma}

\begin{lemma}[{{Pi}} is a category]
  We have as objects the types {{b}} and as morphisms equivalence
  classes of combinators {{b1<->b2}}. One can check:
  \begin{enumerate}
  \item Every object {{b}} has an identity morphism {{b <->b}}.
  \item Composition {{g (o) f}} of morphisms {{f:b1 <->b2}} and
    {{g:b2<->b3}} is given by sequencing {{f (;) g}}. 
  \item Associativity of composition follows from operational
    equivalence of {{f(;)(g(;)h)}} and {{(f(;)g)(;)h}} (where
    {{h:b3<->b4}}). One can check:

%subcode{proof} include main
%@@ f v1 |--> v2
%@ g v2 |--> v3
%@ h v3 |--> v4
%@@ g (;) h  v2 |--> v4
%@@@ f (;) (g (;) h) v1 |--> v4

and

%subcode{proof} include main
%@ f v1 |--> v2
%@ g v2 |--> v3
%@@ f (;) g v1 |--> v3
%@@ h v3 |--> v4
%@@@ (f (;) g) (;) h v1 |--> v4

  \item Composition respects identity: {{id (;) f = f}} and {{g(;)id=g}}.
  \end{enumerate}
\end{lemma}

\begin{lemma}[Symmetric Monoidal (+, 0)]
  {{Pi}} is a symmetric monoidal category with tensor {{+}} and
  monoidal unit {{0}}. For any object {{b1+b2}} the required symmetry
  morphism (called in categorical literature) is
  {{swap+:b1+b2<->b2+b1}}.
\end{lemma}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Small Step Semantics}
